Miniproject - Analytic Number Theory

### 2021-12-16 Analytic number theory/submission.tex
Done my best to answer (vii) in ANT

    Now consider the element $\alpha = \omega_q - xqP \in \mathcal{O}$. Then
    \begin{align*}
        N_{K/\Q}(\alpha) &= \prod_{\sigma: K \to \C \ \text{$\Q$-field homomorphism}} \sigma(\omega_q - xqP) \\
        &= \prod_{\sigma} (\sigma(\omega_q) - xqP) = \mathrm{MiPo}_\Q(\omega_q)(xqP) = \Phi_q(xqP) \neq 1
    \end{align*}
    as $\mathrm{MiPo}_\Q(\omega_q) = \prod_\sigma (\sigma(\omega_q) - X)$.
    Hence, $\alpha$ is not a unit in $\mathcal{O}$.
    On the other hand, $(\alpha)$ is coprime to $(p_i)$ for each $p_i$, as 
    \begin{equation*}
        \omega_q = \alpha - xqP \in (\alpha) + (p_i) \quad \text{and} \quad \omega_q \in \mathcal{O}^\units
    \end{equation*}
    By our assumption, the only prime ideals in $\mathcal{O}$ are the prime ideal factors of $(p_i)$ and $(p)$ for $p \neq p_i$.
    Thus, the prime ideal factorization of $(\alpha)$ consists only of prime ideals $(p), p \neq p_i$ and it follows that $(\alpha) = (n)$ for some integer $n \geq 2$.
    As $\omega_q$ and $xqP \in \Z$ are $\Q$-linearly independent, we see that $n \divides \omega_q$ and $n \divides xqP$.
    However, the former is a contradiction, as $\omega_q \in \mathcal{O}^\units$ is a unit and no $n \geq 2$ is a unit.
\end{proof}
The book also mentions that the general case can be proven by using L-series in algebraic number fields.
I am pretty sure that those are exactly the functions $L(\chi, s)$ that we defined above.
\begin{corollary}[Question (iii)]
    There are infinitely many primes $p$ with $p \equiv 1 \mod 4$.
\end{corollary}
\begin{proof}
    This is just a special case of Prop.~\ref{prop:primes_1_mod_q}.
\end{proof}
\begin{example}[Question (iii)]
    Using a computer, we can also study the actual frequency of prime numbers $\equiv 1, 3 \mod 4$ among e.g. the first $10^8$ integers.
    This seems to indicate that both numbers are asymptotically equal, which seems natural, given the result of Example~\ref{ex:dirichlet_sum_over_primes}.
    For example, there are 332180 primes $\equiv 1 \mod 4$ and 332398 primes $\equiv 3 \mod 4$ smaller than $10^8$.
    To find these numbers, the following python code was used.
    \lstinputlisting[language = python]{./primes.py}
\end{example}

@@

    By Prop.~\ref{prop:representation_mangoldt_series_congruent_a} we have that for $\Re(s) > 1$
    \begin{equation*}
        \rho_a(s) = \sum_{n \equiv a \mod q} \Lambda(n) n^{-s}
    \end{equation*}
    Partial summation yields that for $\Re(s) > 1$ have
    \begin{align*}
        \rho_a(s) =& \sum_{n \equiv a \mod q} \Lambda(n) n^{-s} \\
        =& \lim_{t \to \infty} \Bigl( \underbrace{ t^{-s} \sum_{n < t, \ n \equiv a \mod q} \Lambda(n) }_{ = t^{-s} \Psi_a(t) = t^{-s} O(t\log t) = o(1)} \Bigr) 
        + s \int_1^\infty \Bigl( \underbrace{\sum_{n < t, \ n \equiv a \mod q} \Lambda(n) }_{ = \Psi_a(t)} \Bigr) t^{-(s + 1)} dt \\
        =& s \int_1^\infty \Psi_a(t) t^{-s - 1} dt = s \int_1^\infty O(t^{1/2}\log t) t^{-s - 1} dt \\
        =& O\left( |s| \int_1^\infty \log(t) t^{-\Re(s) - 1/2} dt \right) \\
        =& O\left( |s| \left( \frac 1 {1/2 - \Re(s)} 1^{1/2 - \Re(s)} + \frac 1 {1/2 - \Re(s)} \int_1^\infty t^{-\Re(s) - 1/2} dt \right) \right) \\
        =& O\left( |s| \frac 1 {(1/2 - \Re(s))^2} \right)
    \end{align*}
    However this function has no pole at $s = 1$, a contradiction to Lemma~\ref{prop:rho_pole_at_one}.
\end{proof}
\begin{example}[Question (vi)]
    For all the nontrivial Dirichlet characters $\chi_1, ..., \chi_4$ defined in Example~\ref{ex:nontrivial_dirichlet_characters_mod_5}, we have
    \begin{equation*}
        L(1, \chi_i) \neq 0
    \end{equation*}
    It follows that there are infinitely many primes $\equiv a \mod 5$, for all $a \perp 5$.
\end{example}
\begin{proof}
    By Prop.~\ref{prop:dirichlet_series_at_one} we know that
    \begin{equation*}
        L(1, \chi_i) = \sum_{n \geq 1} \chi(n) n^{-s}
    \end{equation*}
    For the fifth root of unity $\omega \in \C$ such that $\chi_i(1) = \omega$, we thus find
    \begin{align*}
        L(1, \chi_i) =& \sum_{k \geq 0} \sum_{1 \leq n \leq 4} \frac {\chi(5k + n)} {(5k + n)^s} = \sum_{k \geq 0} \frac {\omega} {(5k + 1)^s} + \frac {\omega^2} {(5k + 2)^s} + \frac {\omega^3} {(5k + 3)^s} + \frac {\omega^4} {(5k + 4)^s} \\
        =& \sum_{k \geq 0} \omega \Bigl( \underbrace{\frac 1 {(5k + 1)^s} - \frac 1 {(5k + 4)^s}}_{> 0} \Bigr) + \omega^2 \Bigl( \underbrace{\frac 1 {(5k + 2)^s} - \frac 1 {(5k + 3)^s}}_{> 0} \Bigr)
    \end{align*}
    Hence we have positive coefficients $a_k, b_k > 0$ with
    \begin{equation*}
        L(1, \chi_i) - \sum_{k \geq 0} a_k \omega + b_k \omega^2 = \omega \sum_{k \geq 0} a_k + b_k \omega
    \end{equation*}
    In particular,
    \begin{equation*}
        \Im\left( \frac {L(1, \chi_i)} {\omega} \right) = \sum_{k \geq 0} b_k \Im(\omega) = \Im(\omega) \underbrace{ \sum_{k \geq 0} b_k }_{> 0}
    \end{equation*}
    Since $\chi_i$ is a nontrivial Dirichlet character, we see that $\Im(\omega) \neq 0$ and so $\Im(L(1, \chi_i)/\omega) \neq 0$, thus $L(1, \chi_i) \neq 0$.
    The claim now follows by Prop.~\ref{prop:infinite_primes_congruent}.
\end{proof}
\begin{remark}[Question (vii)]
    TODO
\end{remark}

>>>>

    Now consider the element $\alpha = \omega_q - xqP \in \mathcal{O}$. Then
    \begin{align*}
        N_{K/\Q}(\alpha) &= \prod_{\sigma: K \to \C \ \text{$\Q$-field homomorphism}} \sigma(\omega_q - xqP) \\
        &= \prod_{\sigma} (\sigma(\omega_q) - xqP) = \mathrm{MiPo}_\Q(\omega_q)(xqP) = \Phi_q(xqP) \neq 1
    \end{align*}
    as $\mathrm{MiPo}_\Q(\omega_q) = \prod_\sigma (\sigma(\omega_q) - X)$.
    Hence, $\alpha$ is not a unit in $\mathcal{O}$.
    On the other hand, $(\alpha)$ is coprime to $(p_i)$ for each $p_i$, as 
    \begin{equation*}
        \omega_q = \alpha - xqP \in (\alpha) + (p_i) \quad \text{and} \quad \omega_q \in \mathcal{O}^\units
    \end{equation*}
    By our assumption, the only prime ideals in $\mathcal{O}$ are the prime ideal factors of $(p_i)$ and $(p)$ for $p \neq p_i$.
    Thus, the prime ideal factorization of $(\alpha)$ consists only of prime ideals $(p), p \neq p_i$ and it follows that $(\alpha) = (n)$ for some integer $n \geq 2$.
    As $\omega_q$ and $xqP \in \Z$ are $\Q$-linearly independent, we see that $n \divides \omega_q$ and $n \divides xqP$.
    However, the former is a contradiction, as $\omega_q \in \mathcal{O}^\units$ is a unit and no $n \geq 2$ is a unit.
\end{proof}
In a later chapter, the book also proves the general result \cite[Chapter VII, Theorem 5.14]{neukirch}.
I am pretty sure that those are exactly the functions $L(\chi, s)$ that we defined above.
\begin{corollary}[Question (iii)]
    There are infinitely many primes $p$ with $p \equiv 1 \mod 4$.
\end{corollary}
\begin{proof}
    This is just a special case of Prop.~\ref{prop:primes_1_mod_q}.
\end{proof}
\begin{example}[Question (iii)]
    Using a computer, we can also study the actual frequency of prime numbers $\equiv 1, 3 \mod 4$ among e.g. the first $10^8$ integers.
    This seems to indicate that both numbers are asymptotically equal, which seems natural, given the result of Example~\ref{ex:dirichlet_sum_over_primes}.
    For example, there are 332180 primes $\equiv 1 \mod 4$ and 332398 primes $\equiv 3 \mod 4$ smaller than $10^8$.
    To find these numbers, the following python code was used.
    \lstinputlisting[language = python]{./primes.py}
\end{example}

@@

    By Prop.~\ref{prop:representation_mangoldt_series_congruent_a} we have that for $\Re(s) > 1$
    \begin{equation*}
        \rho_a(s) = \sum_{n \equiv a \mod q} \Lambda(n) n^{-s}
    \end{equation*}
    Partial summation yields that for $\Re(s) > 1$ have
    \begin{align*}
        \rho_a(s) =& \sum_{n \equiv a \mod q} \Lambda(n) n^{-s} \\
        =& \lim_{t \to \infty} \Bigl( \underbrace{ t^{-s} \sum_{n < t, \ n \equiv a \mod q} \Lambda(n) }_{ = t^{-s} \Psi_a(t) = t^{-s} O(t\log t) = o(1)} \Bigr) 
        + s \int_1^\infty \Bigl( \underbrace{\sum_{n < t, \ n \equiv a \mod q} \Lambda(n) }_{ = \Psi_a(t)} \Bigr) t^{-(s + 1)} dt \\
        =& s \int_1^\infty \Psi_a(t) t^{-s - 1} dt = s \int_1^\infty O(t^{1/2}\log t) t^{-s - 1} dt \\
        =& O\left( |s| \int_1^\infty \log(t) t^{-\Re(s) - 1/2} dt \right) \\
        =& O\left( |s| \left( \frac 1 {1/2 - \Re(s)} 1^{1/2 - \Re(s)} + \frac 1 {1/2 - \Re(s)} \int_1^\infty t^{-\Re(s) - 1/2} dt \right) \right) \\
        =& O\left( |s| \frac 1 {(1/2 - \Re(s))^2} \right)
    \end{align*}
    However this function has no pole at $s = 1$, a contradiction to Lemma~\ref{prop:rho_pole_at_one}.
\end{proof}
\begin{example}[Question (vi)]
    For all the nontrivial Dirichlet characters $\chi_1, ..., \chi_4$ defined in Example~\ref{ex:nontrivial_dirichlet_characters_mod_5}, we have
    \begin{equation*}
        L(1, \chi_i) \neq 0
    \end{equation*}
    It follows that there are infinitely many primes $\equiv a \mod 5$, for all $a \perp 5$.
\end{example}
\begin{proof}
    By Prop.~\ref{prop:dirichlet_series_at_one} we know that
    \begin{equation*}
        L(1, \chi_i) = \sum_{n \geq 1} \chi(n) n^{-s}
    \end{equation*}
    For the fifth root of unity $\omega \in \C$ such that $\chi_i(1) = \omega$, we thus find
    \begin{align*}
        L(1, \chi_i) =& \sum_{k \geq 0} \sum_{1 \leq n \leq 4} \frac {\chi(5k + n)} {(5k + n)^s} = \sum_{k \geq 0} \frac {\omega} {(5k + 1)^s} + \frac {\omega^2} {(5k + 2)^s} + \frac {\omega^3} {(5k + 3)^s} + \frac {\omega^4} {(5k + 4)^s} \\
        =& \sum_{k \geq 0} \omega \Bigl( \underbrace{\frac 1 {(5k + 1)^s} - \frac 1 {(5k + 4)^s}}_{> 0} \Bigr) + \omega^2 \Bigl( \underbrace{\frac 1 {(5k + 2)^s} - \frac 1 {(5k + 3)^s}}_{> 0} \Bigr)
    \end{align*}
    Hence we have positive coefficients $a_k, b_k > 0$ with
    \begin{equation*}
        L(1, \chi_i) - \sum_{k \geq 0} a_k \omega + b_k \omega^2 = \omega \sum_{k \geq 0} a_k + b_k \omega
    \end{equation*}
    In particular,
    \begin{equation*}
        \Im\left( \frac {L(1, \chi_i)} {\omega} \right) = \sum_{k \geq 0} b_k \Im(\omega) = \Im(\omega) \underbrace{ \sum_{k \geq 0} b_k }_{> 0}
    \end{equation*}
    Since $\chi_i$ is a nontrivial Dirichlet character, we see that $\Im(\omega) \neq 0$ and so $\Im(L(1, \chi_i)/\omega) \neq 0$, thus $L(1, \chi_i) \neq 0$.
    The claim now follows by Prop.~\ref{prop:infinite_primes_congruent}.
\end{proof}
Finally, we want to reflect on the general approach and highlight how we used of Dirichlet characters.
\begin{remark}[Question (vii)]
    Roughly speaking, both the approach used in the lecture and in this Miniproject consists of three steps.
    \begin{itemize}
        \item Relate the frequency of considered primes to the poles of a Dirichlet series, using partial summation or Perron's formula.

        \item Express the above series as logarithmic derivative of a second function.
        This then relates poles to zeros, by the properties of the Taylor series expansion of meromorphic functions.

        \item Investigate the zeros of the second function.
    \end{itemize}

    It seems like the magic happens mostly in the second step.
    The crucial part here is the connection between multiplicative arithmetic and Dirichlet series.
    In particular, this gives us nice identities like the Euler product.
    However, this will be completely destroyed when we consider functions that are not multiplicative at all.

    More concretely, consider
    \begin{equation*}
        f: \N \to \C, \quad n \mapsto \begin{cases}
            1 & \text{if $n \equiv a \mod q$} \\
            0 & \text{otherwise}
        \end{cases}
    \end{equation*}
    Note that this function is not multiplicative.
    After applying the first step, we are interested in the poles of the meromorphic function
    \begin{equation*}
        \sum_{n \geq 1} f(n) \Lambda(n) n^{-s}
    \end{equation*}
    I will not prove this here, but via partial summation, we can find a meromorphic continuation to $\Re(s) > 0$.

    However, here we get stuck.
    It might seem natural to consider the function
    \begin{equation*}
        \zeta_a(s) := \sum_{n \geq 1} f(n) n^{-s}
    \end{equation*}
    which also has a meromorphic continuation to $\Re(s) > 0$.
    Its derivative is clearly given by
    \begin{equation*}
        \zeta_a'(s) = \sum_{n \geq 1} f(n) \log(n) n^{-s}
    \end{equation*}
    To continue, we might try to consider
    \begin{equation*}
        \zeta_a'(s)\Bigl( \sum_{n \geq 1} \mu(n) f(n) n^{-s} \Bigr) = \sum_{n \geq 1} \frac {(f\log * f\mu)(n)} {n^s}
    \end{equation*}
    But since $f$ has no nice compatibility with multiplication or $*$, it is not easy to make something of this.
    In particular
    \begin{equation*}
        f\Lambda \neq f\log * f\mu \quad \text{and} \quad \frac 1 {\zeta_a(s)} \neq \sum_{n \geq 1} f(n) \mu(n) n^{-s}
    \end{equation*}
    Note that we find $n \equiv a \mod q$ such that $n$ has no nontrivial factor $\equiv a \mod q$.
    E.g. take $n = 9 \equiv 4 \mod 5$, then
    \begin{equation*}
        \log(3) = (f\Lambda)(6) \neq \sum_{cd = 9} f(c)\log(c)f(d)\mu(d) = 0
    \end{equation*}
    Furthermore, the equality
    \begin{equation*}
        \frac 1 {\zeta_a(s)} = \sum_{n \geq 1} f(n) \mu(n) n^{-s}
    \end{equation*}
    would imply
    \begin{equation*}
        1 = \zeta_a(s) \sum_{n \geq 1} f(n) \mu(n) n^{-s} = \sum_{n \geq 1} \frac {(f * f\mu)(n)} {n^s}
    \end{equation*}
    which in turns implies that $f * f\mu = \delta$.
    However, we clearly find counterexamples for that, e.g. for $q = 5, a = 4$ find $(f * f\mu)(1) = 0$.
    The underlying reason is that the divisors of $n$ may lie in a different congruence class modulo $q$ than $n$.

    In some sense, this whole miniproject was just about using the decomposition
    \begin{equation*}
        f = \frac 1 {q - 1} \sum_{\chi \in \mathrm{Dir}(q)} \overline{\chi(a)} \chi
    \end{equation*}
    as then the Dirichlet series become Dirichlet series over $\chi$ instead of $f$, which is much better manageable.
\end{remark}
